---
title: Import WGBS data in a BSseq object
author:
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Import WGBS data produced by Bismark into a BSSeq object.
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2
---

# Prior work

Raw read pairs were adapter- and quality-trimmed using _Trim Galore!_
as follows:

```
parallel -j $threads --xapply trim_galore \
    --paired \
    --output_dir $outdir/$batch ::: $fastq1s ::: $fastq2s
```

Trimmed read pairs were aligned using _Bismark_ as follows:

```
bismark \
    $genome \
    --bowtie2 \
    --nucleotide_coverage \
		--output_dir $outdir/$batch \
		-p $threads \
		--temp_dir $tempdir \
		-1 $paired1s -2 $paired2s
```

Aligned read pairs were deduplicated using _Bismark_ as follows:

```
parallel -j $threads --xapply deduplicate_bismark \
    --bam -p ::: $bams_paired
```

Methylation calls were extracted from deduplicated aligned read pairs
using as follows:

```
parallel -j $threads --xapply bismark_methylation_extractor \
    --output $outdir/$batch
    --bedGraph \
		--buffer_size 10G \
		--scaffolds \
		--cytosine_report \
		--gzip \
		--genome_folder $genomeFolder \
		--no_overlap \
		--paired-end ::: $bams_paired
```

# Settings

Required packages:

```{r pkgs}
library(bsseq)
library(BiocParallel)
library(data.table)
library(GenomicRanges)
library(reshape2)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set up some paths and values relevant for the analysis:

```{r paths}
cpgFolder <- "extract_refined/Merged"
cpgFilePattern <- "CpG_report.txt.gz$"
CPUs <- 2
outDir <- file.path(workdir, "bsseq")
```

Let us created folders as necessary:

```{r dirCreate}
if (!dir.exists(outDir)){
  dir.create(outDir)
}
```

# Prepare phenotype information

First, let us obtain the list of Bismark reports for CpG sites:

```{r cpgFiles}
cpgFiles <- list.files(file.path(workdir, cpgFolder), cpgFilePattern)
length(cpgFiles)
```

From the file names, let us now assemble phenotype information:

```{r pdata}
pdata <- data.frame(
  Sample = sapply(X = strsplit(cpgFiles, '_'), FUN = "[[", 1),
  Infection = factor(
    x = c("Control", "M. bovis")[
      as.numeric(factor(
        x = sapply(
          X = sapply(X = strsplit(x = cpgFiles, split = '_'), FUN = "[[", 1),
          FUN = "substr", start = 1, stop = 1),
        levels = c("C", "M")))],
    levels = c("Control", "M. bovis")),
  Animal = factor(sapply(
    X = sapply(X = strsplit(x = cpgFiles, split = '_'), FUN = "[[", 1),
    FUN = "gsub",
    pattern = "[CM]",
    replacement = ""),
    levels = sort(as.numeric(unique(
      sapply(
        X = sapply(X = strsplit(x = cpgFiles, split = '_'), FUN = "[[", 1),
        FUN = "gsub",
        pattern = "[CM]",
        replacement = ""))))),
  Filename = cpgFiles,
  stringsAsFactors = FALSE
)
dimnames(pdata)
```

We may examine the resulting `data.frame`:

```{r pdata2pandoc, results='asis', echo=FALSE}
pdata2 <- pdata
pdata2$Filename <- paste(
  substr(pdata2$Filename, 1, 10),
  substr(pdata2$Filename, nchar(pdata2$Filename)-16, nchar(pdata2$Filename)),
  sep = "...")
pander::pandoc.table(pdata2)
rm(pdata2)
```

```{r writePdata}
write.csv(pdata, file.path(outDir, 'phenodata.csv'), row.names = FALSE)
```

# Import methylation calls

Let us write a function that parses a gzip-compressed report file
(using the efficient `r CRANpkg("data.table")` `fread` method)
and return a `r Biocpkg("bsseq")` `BSseq` object:

```{r readCpgReportGzip, message=FALSE, eval=FALSE}
read.CpGreport.gz <- function(
  index, pData, folder = ".", sample = "Sample", file = "Filename")
  {
  fread_cmd <- paste("gunzip -c", file.path(folder, pData[index, file]))
  message(pData[index, sample], " : ", fread_cmd)
  dat <- as.data.frame(fread(fread_cmd))
  gr <- GRanges(
    seqnames = dat[,1],
    ranges = IRanges(start = dat[,2], end = dat[,2]),
    strand = dat[,3])
  # Methylation calls
  M <- as.matrix(dat[,4], drop=FALSE)
  # Total calls
  Cov <- as.matrix(dat[,4] + dat[,5], drop=FALSE)
  BSseq(
    M = M,
    Cov = Cov,
    pData = pData[index,],
    sampleNames = pData[index, sample],
    gr = gr)
}
```

We may then import methylation calls for each sample and :

```{r bplapplyImport, eval=FALSE}
BS.list <- bplapply(
  1:nrow(pdata), 
  read.CpGreport.gz,
  pData = pdata,
  folder = cpgFolder,
  BPPARAM = MulticoreParam(CPUs))
```

Let us then combine the individual samples into a single `BSseq` object:

```{r combineList, eval=FALSE}
BS.combined <- combineList(BS.list)
```

We may then free the memory used to store the individual data sets,
to reduce memory usage:

```{r rmBSlist, eval=FALSE}
rm(BS.list, read.CpGreport.gz)
```

Given the signficant time required to parse the data files and assemble the
`BSseq` object, let us save the serialised object to file for faster access
in susbsequent analyses:

```{r saveBScombined, eval=FALSE}
saveRDS(BS.combined, file.path(outDir, 'BS.combined.rds'))
```

```{r importBScombined, echo=FALSE}
BS.combined <- readRDS(file.path(outDir, 'BS.combined.rds'))
```

# Raw statistics

Let us examine the mean coverage of the stranded methylation calls imported
from the raw data files:

```{r covStranded}
meanCov.stranded <- getCoverage(BS.combined, what = "perRegionAverage")
meanCov <- data.frame(Stranded = meanCov.stranded)
summary(meanCov)
write.csv(meanCov, file.path(outDir, "meanCov.csv"))
rm(meanCov.stranded)
```

Let us find the maximum coverage for any given CpG and sample:

```{r baseCoverageStranded}
max(getCoverage(BS.combined))
```

Let us define appropriate cut-offs of coverage:

```{r minCovsStranded}
minCovs <- c(1, 2, 5, 10 , 20, 50, 100, 200, 500, 1000)
```

We may then examine the number of loci found at those levels of coverage
within each sample:

```{r strandedCovs, eval=FALSE}
strandedCovs <-
  sapply(minCovs, function(minCov){
    message(sprintf("Coverage: %i", minCov))
    colSums(getCoverage(BS.combined) > minCov)
  })
colnames(strandedCovs) <- as.character(minCovs)
saveRDS(strandedCovs, file.path(outDir, "strandedCovs.rds"))
```

```{r importStrandedCovs, echo=FALSE}
strandedCovs <- readRDS(file.path(outDir, "strandedCovs.rds"))
```

```{r showStrandedCovs, results='asis'}
pander::pandoc.table(strandedCovs)
```

```{r meltStrandedCovsGG}
strandedCovsGG <- melt(
  strandedCovs,
  varnames = c("Sample", "Coverage"),
  value.name = "Loci")
strandedCovsGG$Group <- ifelse(
  grepl("^C", strandedCovsGG$Sample),
  "Control",
  "M. bovis"
)
```

```{r strandedCovsPlot}
library(ggplot2)
ggplot(strandedCovsGG) +
  geom_line(aes(x = Coverage, y = Loci, colour = Group, group = Sample)) +
  scale_y_log10(
    breaks = 10^(0:9),
    labels = 10^(0:9),
    minor_breaks = rep(1:9, each = 8)*10^(0:8),
    limits = c(1, nrow(BS.combined))) +
  labs(x = "Stranded coverage", y = "Stranded CpG loci")
```

# Unstranded statistics

Let us collapse methylation information from both both strands into a single
value per unstranded CpG locus:

```{r strandCollapse, eval=FALSE}
BS.unstranded <- strandCollapse(BS.combined)
```

```{r BSunstrandedImport}
BS.unstranded <- readRDS(file.path(outDir, 'BS.unstranded.rds'))
```

Let us release memory used to store the large `BSseq` object used to store
stranded methylation calls:

```{r rmBScombined}
rm(BS.combined)
```

Similarly to the previous section, let us collect coverage statistics for
the unstranded methylation calls and append them to the existing summary
table:

```{r covUnstranded}
meanCov.unstranded <- getCoverage(BS.unstranded, what = "perRegionAverage")
meanCov <- cbind(meanCov, data.frame(Unstranded = meanCov.unstranded))
summary(meanCov)
write.csv(meanCov, file.path(outDir, "meanCov.csv"))
rm(meanCov.unstranded)
```

As expected, collapsing methylation in unstranded values naturally doubled the
coverage of each palindromic C~p~G locus.

Again, we may then examine the number of loci found at various levels o
coverage within each sample.

First let us find out the new maximum coverage after collapsing strand
information:

```{r baseCoverageUnstranded}
max(getCoverage(BS.unstranded))
```

Let us add additional cut-offs of coverage, considering the new values:

```{r minCovsUnstranded}
minCovs <- c(minCovs, 1200, 1500)
```

```{r unstrandedCovs, eval=FALSE}
unstrandedCovs <-
  sapply(minCovs, function(minCov){
    message(sprintf("Coverage: %i", minCov))
    colSums(getCoverage(BS.unstranded) > minCov)
  })
colnames(unstrandedCovs) <- as.character(minCovs)
unstrandedCovsGG <- melt(
  unstrandedCovs,
  varnames = c("Sample", "Coverage"),
  value.name = "Loci")
unstrandedCovsGG$Group <- ifelse(
  grepl("^C", unstrandedCovsGG$Sample),
  "Control",
  "M. bovis"
)
saveRDS(unstrandedCovsGG, file.path(outDir, "unstrandedCovs.rds"))
```

```{r importUnstrandedCovsGG}
unstrandedCovsGG <- readRDS(file.path(outDir, "unstrandedCovs.rds"))
```

```{r unstrandedCovsPlot}
library(ggplot2)
ggplot(unstrandedCovsGG) +
  geom_line(aes(x = Coverage, y = Loci, colour = Group, group = Sample)) +
  scale_y_log10(
    breaks = 10^(0:9),
    labels = 10^(0:9),
    minor_breaks = rep(1:9, each = 8)*10^(0:8),
    limits = c(1, nrow(BS.unstranded))
  ) +
  labs(x = "Untranded coverage", y = "Untranded CpG loci")
```

We may also query the aggregated average coverage per experimental group
(sum of average C~p~G coverage across samples within each infection group):

```{r aggregateCoverageUnstranded}
tapply(
  getCoverage(BS.unstranded, what = "perRegionAverage"),
  BS.unstranded$Infection,
  sum
)
```

# C~p~G loci with zero coverage

To reduce further memory usage, we may remove altogether loci without
methylation call in any sample (in the unstranded data set):

```{r BSrmZero}
BS.rmZero <- BS.unstranded[rowSums(getCoverage(BSseq = BS.unstranded)) > 0,]
saveRDS(BS.rmZero, file.path(outDir, 'BS.rmZero.rds'))
```

Let us find out the proportion of loci discarded in this manner:

```{r proportionNotCovered}
1 - (nrow(BS.rmZero) / nrow(BS.unstranded))
```

This marginal data reduction indicates that a very small proportion of
C~p~G loci were not covered by the WGBS.

We may now free again the memory used to store the unstranded data set prior
to filtering:

```{r removeBSunstranded}
rm(BS.unstranded)
```

Let us examine the effect of this marginal filtering on overall coverage
in the data set:

```{r covNoZero}
meanCov.rmZero <- getCoverage(BS.rmZero, what = "perRegionAverage")
meanCov <- cbind(meanCov, data.frame(NoZero = meanCov.rmZero))
write.csv(meanCov, file.path(outDir, "meanCov.csv"))
summary(meanCov.rmZero)
rm(meanCov.rmZero)
```

# C~p~G loci covered in all samples

Let us examine the coverage of C~p~G loci covered in all samples:

```{r allCovered}
BS.allCovered <- BS.rmZero[
  rowSums(getCoverage(BS.rmZero) > 0) == ncol(BS.rmZero),
  ]
saveRDS(BS.allCovered, file.path(outDir, 'BS.allCovered.rds'))
```

We may then examine the proportion of loci covered in all samples, relative
to the near-complete set of loci covered in at least a single sample:

```{r proportionAllCovered}
nrow(BS.allCovered) / nrow(BS.rmZero)
```

This much more modest proportion suggests that the shallow sequencing of
samples in this data set provides limited information in individual samples,
in contrast to an overall near-complete coverage of the genome across all
samples.

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
